---
title: "Lab 02, Fall 2020"
author: "Your Friendly GEs!"
date: "10/5/2020"
output:
  md_document:
    variant: markdown_github
  html_document:
    df_print: paged
    keep_md: yes
    smart: yes
    theme: cosmo
    toc: yes
    toc_depth: '3'
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
  header-includes: \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Week 2: Introduction to R part 2:

Hello everybody! Welcome to week 2!

## Lesson 0: Quick review from last time

We learned last time about:

- Objects in R
- Functions in R
- Math calculations in R
- Object types in R
- How to load packages
- Dataframes
- Basic Plots
- Regression Models

This time, we will introduce the tidyverse, and show you how to load your own data into the R workspace.

Lab outline:
  
  - Lesson 1: Directories
  - Lesson 2: Loading Data
  - Lesson 3: Data Manipulation
  - Lesson 4: Introduction to ggplot2

## Lesson 1: Directories

Before we learn to read data into R, it would be helpful to know how to tell R
where it is. This is important because later in the class we need to load files from our local machine. Eventually, we want to start using our own data instead of data contained in a package. 
 
 *Note: Windows and R can both use directory separator `/`. Windows can also use `\\` (must be double).*
 
 - To find your current directory: `getwd()`. "wd" stands for working directory.

```r
getwd()
```

 
 - To change directories: `setwd()`
 

```r
setwd("..")
getwd()
```
 
 - To show files in the current directory: `dir()`
 

```r
dir()
```
 
 - Notice that directories are characters (surrounded by `""`).
 - RStudio will help you complete file paths when you hit tab while typing in the console.
 - You can save an object with the directory location and then return to that directory using `setwd()`.
 

Example:
 
```r
my_dir <- "C:/Users/Ajdic/OneDrive - University Of Oregon/GE/EC 421 - Q12020/Lab_02"
setwd(my_dir)
```
 

## Lesson 2: Loading data

R can read in data from just about any source/format. Today we're going to cover reading data saved in CSVs (comma-separated variables).

First, we'll load the `tidyverse` package, which will actually load several packages (we want `readr`). The `base` (basic) installation of R already has a function for reading CSVs, but the function in `tidyverse` (`readr`) is a bit nicer. 

Recall our workflow packages from last time:
```{R, load tidyverse}
library(pacman)
p_load(tidyverse)
```

You can get to the help files in R/RStudio using `?`. Let's check out the help file for `read_csv`.
```{R, help read_csv}
?read_csv
```


Next we are going to work with a `.csv` to practice. So let's go get one online!

The data we are going to work with in this lab with be a publically available data set that tracks monthly U.S. Presidential approval ratings from 2001 to 2006. We can find, download, and load it into our meomory with the following procedure:

- Step 1: Find the `.csv` from the following link - https://www.stat.auckland.ac.nz/~wild/data/Rdatasets/ 
  - It is titled 'approval' - 'U.S. Presidential Approval Data', use Ctrl + F or Command + F to search the website
- Step 2: Download the file to a folder on your computer, note the directory.
  - You can find the file's filepath by either (windows) right clicking on the file and looking in **properties**, under location
  - Or, (Mac) right click on the file, hold down alt and select the **copy as filepath** option

*One of the most important habits to have when coding is to be extremely organized. Think hard about where you save things, it will save you time in the future if you are smart about it.*

- Step 3: Navigate R to the folder loacation, using the `setwd(my_dir)` command, setting my_dir to the filepath

- Step 4: Read the data into R, using `read_csv("./approval.csv")`


```{R}
read_csv("./data/approval.csv") %>% head()
```

*Recall from last time - using `head()` is a good way to get a visual representation of the data set when first *


- Notice that we read the data, but it just printed to screen. We want to assign the data to an object (give it a name). `pres_df <- read_csv("presidents.csv")`


```{R}
pres_df <- read_csv("./data/approval.csv")
```

Additionally, `glimpse()` and `summary()` are two of my favorite functions when working with a new data set

```{R}
glimpse(pres_df)
pres_df %>% summary()
```

Most of the varaibles are self explanatory. However, a few are not super clear

- 'sept.oct.2001': A binary variable for 9-11-2001      (why would this be included??)
- 'iraq.war': A binary variable for the beginning of the Iraq war
- 'avg.price': Average price of a barrel of oil



**Question: What does the following code do? Try to figure this out before running**

```{r}
head(table(pres_df$year > 2003, pres_df$approve > 50))
```



**What about this code?**

```{r}
pres_df %>% filter(year > 2003, approve > 50)
```


One other way to quickly summarize a new dataframe/variable is to create a visual. The `plot()` function is a simple, easy to use function to make quick graphs
```{r}
plot(pres_df$year, pres_df$approve, pch = 19, xlab = "Bad Time", ylab = "Approval Rate %")
```

But wait, this looks weird. We are only plotting over years, not months and We have more information in the 'month' variable. How can we make this plot more informative?

Let's make our time variables flexible. We'll do this with the lubridate package. 

```{r}
p_load(lubridate)
```


We have two time/date related variables: 'year' and 'month'. But what if we wanted to have one time/date variable that includes both the year and month varaible? Let's just make one!


We can do this using another dplyr function: `mutate()`

*Mutate is your best friend*


What is mutate? Mutate is a function that writes new variables.

```{r}
#?mutate()
```


Here are a few examples:

```{r}
pres_df %>% mutate(eleven = 11) #creates a variable for each observation with the value of 11
pres_df %>% mutate(avg.price_levels = (avg.price/max(avg.price))) # This creates a new varaible for 'avg.price' but normalizes the maximum to 1
```

Notice, the variables are not saved to your dataframe.
```{r}
head(pres_df)
```

In order to change you dataframe, you must assign the mutated dataframe to an object or your previous dataframe object. For example:

```{r}
pres_df_example <- pres_df %>% mutate(eleven = 11)
head(pres_df_example)
```

Now lets try to use 'year' and 'month' to make a 'year_month' varaible using the lubridate package

```{r}
pres_df <- pres_df %>% mutate(year_month = make_datetime(year, month))
head(pres_df)
tail(pres_df)
```


Now, using this new variable, we can create informative plots over time. 

```{r}
plot(pres_df$year_month, pres_df$approve, pch = 19, xlab = "Time", ylab = "Approval Rating")
```


There, this looks much better.



## Lesson 3: Data manipulation

You will often need to do some data work before you're ready to actually implement econometric/statistical modeling. This data work generally includes adding new variables, dropping unwanted observations, and merging datasets.

In R, there are many ways to do the same task. For the most part, we're going to try to stick to the functions contained in the `tidyverse`.

The three big `tidyverse` functions that we will go over today are:

- `mutate()`
- `select()`
- `filter()`

We have already used `mutate()` above in the previous lesson but let's go over it again because it is very important


### Creating (mutating) new variables

`dplyr` also makes it easy to create add new variables using `mutate`. Mutate takes as arguments a dataframe and then some function to use to create a new column. Mutate will then output an **entirely new dataframe** with your new column added on.

Let's work through a few examples:

A column of ones: 

```{r}
pres_df %>% mutate(ones = 1)
```


If we wanted to keep this variable to use later, we would need to re-assign the output of `mutate` to `pres_df` (or a brand new dataframe) to save it.

Also note that we can give `mutate()` a single value and it will fill in the whole column (variable) with that value. This can be handy.

We can make a time trend as well. I know that the data is 65 observations long, but let's just say I didn't know that. Do you remember how we could figure it out?


```{r}
nrow(pres_df)
```


Now, we can make a timetrend using this command directly


```{r}
#This creates a time trend
pres_df <- mutate(pres_df, time_trend = 1:65)
#This command is equivalent to
pres_df <- mutate(pres_df, time_trend = 1:nrow(pres_df))
#This command, however, lets us change the number of observations in our dataframe without changing our code for the time trend.
```

We can also easily create dummy variables. Let's create a dummy variable for "post 9/11" (observation number 8 in the time trend): 

```{r}
pres_df <- mutate(pres_df, post_911 = time_trend >= 8)
```

However, what kind of variable is 'post_911'? Let's look.


```{r}
class(pres_df$post_911)
```

Recall from last lecture that this means it is either equal to `TRUE`, `FALSE` or `NA`. We might want to have a more traditional, 0,1 dummy variable. We can do this with the following command:


```{r}
#Use as.integer to cast a true-false value to 0,1.
pres_df <- mutate(pres_df, post_911 = as.integer(time_trend >= 8))
pres_df$post_911
```

If we want to convert it back, we can simply do so using the `as.logical` command


```{r}
pres_df$post_911 <- as.logical(pres_df$post_911)
```


Let's create another dummy variable for "post beginning of Iraq war" (starts with observation number 26 in the time trend): 


```{r}
pres_df <- mutate(pres_df, post_iraq = time_trend >= 26)
#now, our dataframe pres_df has a new column called "post_iraq"
pres_df$post_iraq
```



### Selecting columns
Last time, we covered two different methods of selecting a dataframe's columns. The first was by column name

```{r}
head(pres_df$year)
```


The second was by index number


```{r}
#Year is the third column in our pres_df dataframe
head(pres_df[,3])
```


The `dplyr` package however (part of the `tidyverse`) offers a nice alternative (plus a lot more functionality).

Let's start with the `select` function. We can grab a column with its name using select


```{r}
#Select(dataframe_name, column_name)
pres_df %>% select(year) %>% head()
```

 Alternatively, we can use the pipe operator `%>%`. This takes the dataframe on the left side and passes it to the right. This lets us pass one command to another.
 
 Let's filter our dataframe down to only the `year` and `month` variables, and then pass that dataframe to another filter command that will select only the `month` variable.
 
```{r}
#first, select month and year, then, select month
pres_df %>%
  select(month) %>%
  head()
```



### Filtering observations

We already know of some ways to do this, either by mentioning the name of a variable, or by calling a row/column by index.

Grab some of the observations from a data frame:
 - Grab the first 4 rows: `pres_df[1:4,]`
 - 6th, 8th, 12th, 13rd rows: `pres_df[c(6, 8, 12, 13),]`
 

```{r}
head(pres_df[1:4,])
```


```{r}
head(pres_df[c(6, 8, 12, 13),])
```

 
 `dplyr` allows us to get a bit more sophisticated:
We can now filter data using a single variable, using the `filter()` command

Only observations in 2001:

```{r}
filter(pres_df, year == 2001)
```

Only observations in 2001 OR 2002:

```{r}
tail(filter(pres_df, year %in% c(2001, 2002)))

# Using pipes

pres_df %>%
  filter(year %in% c(2001, 2002)) %>%
  tail()
```

Only observations between 2001 and 2003: 

```{r}
pres_df %>% filter(between(year, 2001, 2003)) %>% head()
```



Filter observations using multiple variables
`sept.oct.2001` is `1` OR `iraq.war` is `1`. Recall our 'or' symbol, |

```{r}
filter(pres_df, sept.oct.2001 == 1 | iraq.war == 1)
```



## Lesson 04: Introduction to ggplot2

Finally lets quickly go over the most powerful visualization tool for this class: `ggplot()`. This function deserves a lesson by itself. You can create a lot of pretty graphs and animations using ggplot and we will just scratch the surface today.

To install the `ggplot2` package run the following code in your console: `install.packages('ggplot2')`

```{r}
p_load(ggplot2)
```

Now lets recreate the graph from the earlier lesson using ggplot

```{r}
ggplot(pres_df, aes(x = year_month, y = approve)) +
  geom_line() +
  ggtitle('Presidential Approval Rating') +
  xlab('Time')
```

We can add some other options that are helpful to visualize the presidential data.
For example, we can add a vertical line to indicate the time of 9.11 on our graph to see the change of the approval rate after the event. 

```{r}
ggplot(pres_df, aes(x = year_month, y = approve)) +
  geom_line(show.legend = TRUE) +
  ggtitle('Presidential Approval Rating') +
  xlab('Time')  +
  geom_vline(xintercept =	as.POSIXct("2001-09-01"), col="red")
```

## Some questions being asked from the previous lab
 
- What does `%>%` sign do exactly and why do we need it?

As mentioned earlier, this is called a pipe operator `%>%`. This operator would forward a value or the result of the expression on the left to the next function call. This lets us pass one command to another. The reason we need this operator is to make our codes more efficient and legible. 

For example, the following two commnads return exactly the same result. However, notice that the first one is a "nested" option which the functions are nested within one another. The problem is that if you add more functions using nested option, it's hard to keep track of what exactly the code is doing and becomes easier to make mistakes when making updates to your codes. 

Thus the alternative option is to use a pipe operator, which makes our codes more readable and easier to spot a mistake if the r gives us an error. 
```{r}
# Nested version
tail(filter(pres_df, year %in% c(2001, 2002)))

# Using pipes
pres_df %>%
  filter(year %in% c(2001, 2002)) %>%
  tail()
```

- What does `%in%` do? 

`%in%` operator is used to identify if the left hand side vector is included in the right hand side vector. Let's see the examples below. 

```{r}
6:10 %in% 1:36
```
As you can see, the sequence from 6 to 10 is included in sequence from 1 to 36. Therefore, r is going to return 5 logical values of TRUE. Let's see another example.

```{r}
6:10 %in% 1:8
```

As you can see, some elements in 6:10 sequence are included in 1:8 and some elements are not. Here, since 6, 7, 8 are included in the right hand side sequence while the rest of the elements are not,  we are going to observe three logical value of TRUE and 2 logical value of FALSE. 

Back to our example, the following command will return 65 logical values consisting of either TRUE or FALSE. It's going to return TRUE if an observation falls either in 2001 OR 2002. It's going to return FALSE if an observation falls in other years.  

```{r}
pres_df$year %in% c(2001, 2002)
```

Now, if we wrap around it with filter function, it's going to select only the observations with "TRUE" out of the total 65 observations. From that filtered dataset, tail function would make you able to view the last 6 observations.  

```{r}
pres_df %>%
  filter(year %in% c(2001, 2002)) %>%
  tail()
```
